---
title: "Getting Started with rjuliabugs: A Step-by-Step Logistic Regression Example with Random Effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{walkthrough_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning=FALSE,message=FALSE,
  comment = "#>"
)
```

This document provides a complete walkthrough for fitting a simple logistic regression model with random effects using the `rjuliabugs` package. Rather than focusing on deep statistical analysis, the main goal of this guide is to demonstrate how to use the functionality offered by `rjuliabugs` to fit BUGS models from R by leveraging modern probabilistic programming tools available in Julia.

In particular, `rjuliabugs` allows users to fit models using advanced samplers such as Hamiltonian Monte Carlo (HMC) and automatic differentiation, as implemented in `JuliaBUGS`. This integration brings the power and flexibility of modern Bayesian computation into the R ecosystem with full compatibility of well know packages as `bayesplot`, `coda`, `posterior` which provide nice and neat visualizations and diagnosticis for a full Bayesian workflow.

For more technical details about `JuliaBUGS`, please refer to [this link](https://github.com/TuringLang/JuliaBUGS.jl). To learn more about probabilistic programming in Julia, we also recommend the [Turing.jl](https://turinglang.org/) project.

## Prerequisites

This vignette assumes that `rjuliabugs` has already been correctly installed and set up. The goal here is to provide a working example under the assumption that your installation is functional.

For detailed installation instructions and troubleshooting information, please refer to the [README](https://github.com/MateusMaiaDS/rjuliabugs/). The installation guide includes the following sections:

- [Installing Julia](#installing-julia)
- [Troubleshooting JuliaCall setup](#troubleshooting-juliacall-setup)
  - [Julia not found](#error-type-julia-not-found)
  - [R not found](#error-type-r_home-not-found)
  - [Other installation issues](#other-installation-issues)
- [Using `rjuliabugs` in parallel](#using-rjuliabugs-in-parallel)


## Data Preparation

To demonstrate a first example, we will use the same dataset presented in the official `JuliaBUGS` illustration. The case concerns the proportion of seeds that germinated on each of 21 plates. To work with this data, we need to create a named list in R that contains all the variables required by the model.


```{r data_list}
data <- list(
    r = c(10, 23, 23, 26, 17, 5, 53, 55, 32, 46, 10, 8, 10, 8, 23, 0, 3, 22, 15, 32, 3),
    n = c(39, 62, 81, 51, 39, 6, 74, 72, 51, 79, 13, 16, 30, 28, 45, 4, 12, 41, 30, 51, 7),
    x1 = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1),
    x2 = c(0, 0, 0, 0, 0, 1, 1, 1, 1, 1, 1, 0, 0, 0, 0, 0, 1, 1, 1, 1, 1),
    N = 21
)
```


Users already familiar with fitting models using `JAGS` or `Stan` will recognize the structure of passing a named list as the data input for the model. Here, we follow the same convention.

One important disclaimer concerns the definition and use of numeric values, particularly integers. When specifying an integer in the data list, make sure it is not written with a decimal point. For example, if a variable like `N` (e.g., the total number of plates) is defined as `21.0` instead of `21`, `JuliaCall` will automatically convert it as a `Float`, which may lead to an error if the model expects an `Integer`. Always use strict integer values when required by the model.

Inspecting closely each element we have `r[i]` as the number of germinated seeds and `n[i]` as the total number of seeds on the $i$-th plate. Let $p_{i}$ be the probability of germination on the $i$-th plate. Then, the model is defined by:


$$
\begin{aligned}
\text{logit}(p_i) &= \alpha_0 + \alpha_1 x_{1i} + \alpha_2 x_{2i} + \alpha_{12} x_{1i} x_{2i} + b_i \\
b_i &\sim \mathcal{N}(0, \tau) \\
r_i &\sim \text{Binomial}(p_i, n_i)
\end{aligned}
$$

where $x_{1i}$ and $x_{2i}$ are the seed type and root extract of the $i$-th plate. 

## BUGS model

Once we have our data prepared, we can define the model using the original BUGS syntax. In `rjuliabugs`, we write the model as a string in R, following the same structure used in other R packages that interface with BUGS-like tools:

```{r juliabugs_code}
model_def <- "
model {
    for (i in 1:N) {
        r[i] ~ dbin(p[i], n[i])
        b[i] ~ dnorm(0.0, tau)
        logit(p[i]) <- alpha0 + alpha1 * x1[i] + alpha2 * x2[i] +
                       alpha12 * x1[i] * x2[i] + b[i]
    }
    alpha0 ~ dnorm(0.0, 1.0E-6)
    alpha1 ~ dnorm(0.0, 1.0E-6)
    alpha2 ~ dnorm(0.0, 1.0E-6)
    alpha12 ~ dnorm(0.0, 1.0E-6)
    tau ~ dgamma(0.001, 0.001)
    sigma <- 1 / sqrt(tau)
}
"
```

Users familiar with R packages like `R2jags`, `rjags`, or `R2WinBUGS` will recognize this approach: the model is defined as a string using the BUGS modeling language. For users coming from Stan (e.g., via `rstan` or `cmdstanr`), the idea is similar: the model is written as a probabilistic program inside a string object, then compiled and executed by the backend.

This syntactic similarity is not accidental. It is a design choice of `rjuliabugs`, which aims to allow users to reuse models originally written for `JAGS` or `WinBUGS`, while taking advantage of modern features available in Julia—such as Hamiltonian Monte Carlo and automatic differentiation—via `JuliaBUGS`.

We also remind users that they should be familiar with the `BUGS` syntax, as syntax-related mistakes will lead to errors during model compilation. For additional guidance, see the [Miscellaneous Notes on BUGS](https://turinglang.org/JuliaBUGS.jl/v0.9/BUGS_notes/) from the JuliaBUGS documentation, as well as the [BUGS Developer Manual](https://turinglang.org/JuliaBUGS.jl/v0.9/BUGS_notes/).

For example, writing `1.0E-6.0` instead of `1.0E-6` is a syntax error in BUGS, even though the numerical meaning is equivalent. These types of issues can be subtle but will cause the model to fail during parsing, so attention to detail when writing BUGS code is important.

## Inference / Running the Sampler

Once the data is defined and the model is set up, we can run the sampler to perform inference. The setup
for the HMC sampler uses Not-U-Turn Sampler (NUTS) with the target acceptance probability
$(\delta=0.8)$ for step size adaptation. This is done using the main function, `juliaBUGS`. For example:

```{r fitting_the_sampler}
library(rjuliabugs)

posterior <- juliaBUGS(
  data = data,
  model_def = model_def,
  params_to_save = c("alpha0", "alpha1", "alpha2", "alpha12", "sigma"),
  n_iter = 2000,
  n_warmup = 1000,
  n_discard = 1000,
  n_chain = 4,
  use_parallel = TRUE,
  n_thin = 1
)
```

This will return the MCMC chains, and a brief inspection can be done through `summary` function

```{r summary}
# Generating the summary of the code
summary(posterior,get_summary = FALSE,get_quantiles = FALSE)
```

If we wish to obtain the summary statistics for each saved parameter, we can set the argument `get_summary = TRUE` when calling `summary()`. Additionally, to include the quantiles associated with the posterior distributions, set `get_quantiles = TRUE`. To inspect only the summary output in the same format as displayed in `JuliaBUGS`, set `julia_summary_only = TRUE`. This will return the native Julia-style summary.

For further details, see the documentation by running `?summary.rjuliabugs`.


## Inspecting the Posterior Samples

Once sampling is done, the `juliaBUGS` function returns an object of class `rjuliabugs`. This object contains:

- `params`  
  Posterior samples, in the format specified by `posterior_type`. These are the parameters defined to be saved in the argument `params_to_save`. The format of this object depends on the prior definition of the argument `posterior_type` when calling `juliaBUGS`. The default is `"array"`, which is compatible with most R libraries that work with posterior samples, and returns a 3D numeric array (iterations × chains × parameters). Other available types are `"rvar"`, `"mcmc"`, and `"draws"`. See `?juliaBUGS` for more details.

- `name`  
  Character string identifying the Julia sampler object. As mentioned before, `rjuliabugs` relies heavily on `JuliaCall` to integrate the R interface with Julia. While using the package, a Julia session runs in the background and generates the `Chains` object containing all posterior samples, which can be accessed through the `name` argument. This name can be defined when calling `juliaBUGS` through the `name` parameter and serves as the key to retrieve the fitted sampler. If `juliaBUGS` is called multiple times with the same name, `rjuliabugs` automatically generates a new unique name to avoid overwriting. To delete a fitted model object from the Julia environment, call `delete_julia_obj()` with the corresponding name. See `?delete_julia_obj` for more details.

- `sampler`  
  Sampler object returned by `AbstractMCMC.sample`. This is a Julia object of type `"JuliaObject"`. For more details, see the documentation of `julia_call()` in the `JuliaCall` package.

- `n_threads`  
  Number of Julia threads detected. This is important to verify whether Julia properly recognized the available threads and if the sampler ran in parallel correctly. If you set `parallel = TRUE` when calling `juliaBUGS` but this value is one, the sampler actually ran sequentially.

- `mcmc`  
  List of MCMC configuration parameters. It includes all MCMC-related arguments passed when calling `juliaBUGS`.

- `control`  
  Control options used in the sampler setup. For more details, see the documentation of `?juliaBUGS`.


## Visualizing Posteriors

Once the posterior samples are obtained, further diagnostic and visualization tools are easily accessible through well-known packages such as `bayesplot` and `posterior`. For a full walkthrough of visualizations and diagnostics, check the other vignette (currently in progress).

For a quick illustration, let's observe the density plot of the posteriors and their traceplots.

```{r bayes_plotposterior, fig.align='center', fig.width=4, fig.height=3, dpi=150}
library(bayesplot)

# Plotting the posterior density for the saved parameters
bayesplot::mcmc_areas(
  posterior$params,
  pars =  c("alpha0", "alpha1", "alpha2", "alpha12", "sigma"),
  prob = 0.8
)
```

For the traceplots, we subset only the $\alpha_{0}$ and the $\sigma$ parameters:

```{r bayes_traceplot, fig.align='center', fig.width=4, fig.height=3, dpi=150}

# Plotting traceplots for alpha0 and sigma
bayesplot::mcmc_trace(
  x = posterior$params,
  pars = c("alpha0", "sigma"),
  n_warmup = 0,
  facet_args = list(nrow = 2)
)
```

Similarly, diagnostics are also available from the posterior package:

```{r diagnostics}
library(posterior)

# For other diagnostics, see the help page
help("diagnostics", "posterior")

# Example: effective sample size
posterior::ess_basic(posterior$params[,,"alpha0"])
```

## Saving and loading the `rjuliabugs` model

Lastly, it is important to be able to save and load a `rjuliabugs` model. To do it so we will use the `save_rjuliabugs()` 
to save the `posterior` object

```{r save,eval=FALSE}
save_rjuliabugs(rjuliabugs_model = posterior,
                file = "rjuliabugs_model.rds",
                chains_file = "rjuliabugs_chains.jls")
```

and, to load we use `load_rjuliabugs()`

```{r load, eval = FALSE}
posterior <- load_rjuliabugs(file = "rjuliabugs_model.rds")
```
